---
title: "CRAZIE Ordination Analyses"
author: "Zechariah Meunier"
date: "Last updated `r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
### Libraries and Data

Load libraries:  

* `ggplot2` for creating data visualizations
* `ggConvexHull` for adding convex hull geom to ggplot2
* `vegan` for NMDS ordinations
* `pairwiseAdonis` for multilevel pairwise PERMANOVA
* `extrafont` for custom fonts
* `rcartocolor` for custom colors
* `grid` for graphical objects
* `gridExtra` for grid graphics
* `tibble` for simple data frames
* `ggrepel` for repelling overlapping text labels
```{r message=FALSE}
library(ggplot2)
#library(ggConvexHull)
library(vegan)
#library(pairwiseAdonis)
library(extrafont)
loadfonts(device = "all", quiet = TRUE) 
library(rcartocolor)
library(grid)
library(gridExtra)
library(tibble)
library(ggrepel)
```

Create color vectors unique to each factor.
```{r}
trtcolors <- c("#377eb8", "#e41a1c", "#4daf4a", "black")

capecolors <-  carto_pal(n = 5, "ag_Sunset")

sitecolors <- c("#542788","#8073ac","#313695",
                "#4575b4","#74add1","#abd9e9",
                "gold4","gold","#fee090","#fdae61",
                "#f46d43","#d73027","#a50026")

periodcolors <- c("dodgerblue","firebrick","#f46d43")
```

Create explanatory variable matrix (`varmat`) by joining sea surface temperature anomaly, upwelling, and Pisaster size data.
```{r}
varmat <- 
  left_join(
    left_join(
      left_join(dscmat_subset, Lag0SSTa, by = c("Cape", "Year")),
    annual_BEUTI, by = c("Cape", "Year")),
  meanSizeAnnual, by = c("Cape", "Year"))

varmat <- dscmat_subset
varmat$Cape <- factor(varmat$Cape, levels = c("CF","CP","CB","CM"))
varmat$Site <- factor(varmat$Site, levels = c("FC","BB","MB","YB","SH","TK","CBN",
                                              "CBS","POH","RP","CMN","CMS","KH"))
varmat$Year <- factor(varmat$Year, levels = c("2006","2007","2008","2009","2010",
                                              "2011","2012","2013","2014","2015",
                                              "2016","2017","2018","2019","2020"))
varmat$Trt <- factor(varmat$Trt, levels = c("C","R","A","I"))
varmat$Rep <- factor(varmat$Rep)

table(varmat$binSSTa, varmat$Year)
plot(varmat$meanSSTa, varmat$meanPisDiam,
     xlab = "Intertidal Water Temperature Anomaly",
     ylab = "Pisaster Diameter")
```


### Nonmetric Multidimensional Sclaing (NMDS)

Perform NMDS in 2 dimensions (k=2) with Bray-Curtis dissimilarities per cape and treatment.
```{r}
NMDScape <- "CF" #specify cape
NMDStrt <- "I" #specify treatment
NMDSmat <- filter(sslFGmat, row.names(sslFGmat) %in% 
                    filter(dscmat_subset, Trt==NMDStrt, Cape==NMDScape)$PlotID)
envmat <- filter(varmat, row.names(sslFGmat) %in% 
                    filter(dscmat_subset, Trt==NMDStrt, Cape==NMDScape)$PlotID)

NMDS <- metaMDS(NMDSmat, distance = "bray", k=2, trace=2, autotransform = FALSE,
                maxit = 300, trymax = 30, halfchange = FALSE)
stressplot(NMDS)
```

Save NMDS scores in a new data frame and add factors to it.
```{r}
scoresmat <- as.data.frame(scores(NMDS)[[1]])
scoresmat$Cape <- factor(dscmat$Cape[match(rownames(scoresmat), dscmat$PlotID)],
                                levels = c("CF","CP","CB","CM"))
scoresmat$Site <- factor(dscmat$Site[match(rownames(scoresmat), dscmat$PlotID)],
                                levels = c("FC","BB","MB","YB","SH","TK","CBN",
                                           "CBS","POH","RP","CMN","CMS","KH"))
scoresmat$Year <- factor(dscmat$Year[match(rownames(scoresmat), dscmat$PlotID)],
                                levels = c("2006","2007","2008","2009","2010",
                                           "2011","2012","2013","2014","2015",
                                           "2016","2017","2018","2019","2020"))
scoresmat$Trt <- factor(dscmat$Trt[match(rownames(scoresmat), dscmat$PlotID)],
                                levels = c("C","R","A","I"))
scoresmat$PeriodMHW <- factor(dscmat$PeriodMHW[match(rownames(scoresmat), dscmat$PlotID)],
                                levels = c("Before","During","After"))
plot(scoresmat)

sppvec <- as.data.frame(scores(NMDS)[[2]])
```

Create basic plot.
```{r}
plot(NMDS, xlim = c(-0.85,0.85), ylim=c(-0.85,0.85))
ordiellipse(scoresmat[,1:2], groups = scoresmat$PeriodMHW, conf = 0.95)
```

Fit functional groups and environmental conditions to the NMDS scores.
```{r}
sppfit <- envfit(NMDS, NMDSmat, permutations = 999)
head(sppfit)
sigspp <- as.data.frame(scores(sppfit, display = "vectors")) %>%
  mutate(Pval = sppfit$vectors$pvals) %>%
  rownames_to_column(var = "FuncGroup") %>%
  mutate(Abbreviation = recode(FuncGroup,
                               "Substrate" = "Rock",
                               "Acorn barnacles" = "Balanoids",
                               "Gooseneck barnacles" = "Goosenecks",
                               "Crustose fleshy algae" = "CFA",
                               "Crustose coralline algae" = "CCA",
                               "Articulated coralline algae" = "ACA",
                               "Bladed algae" = "BA",
                               "Finely branched algae" = "FBA",
                               "Coarsely branched algae" = "CBA",
                               "Turf algae" = "TA",
                               "Kelps & fucoids" = "Kelps")) %>%
  filter(Pval <= 0.001)

envfit <- envfit(NMDS, envmat[,c(17,19,22)], permutations = 999, na.rm = TRUE)
head(envfit)
sigenv <- as.data.frame(scores(envfit, display = "vectors")) %>%
  mutate(Pval = envfit$vectors$pvals) %>%
  rownames_to_column(var = "Factor") %>%
  mutate(Abbreviation = recode(Factor,
                               "meanSSTa" = "SSTa",
                               "meanPisDiam" = "Pisaster\u2205",
                               "meanBEUTI" = "BEUTI"
                               )) %>%
  filter(Pval <= 0.05)
```

Calculate centroids for each factor based on NMDS scores.
```{r}
trtcent <- aggregate(cbind(NMDS1, NMDS2) ~ Trt, data = scoresmat, FUN = "mean")
capecent <- aggregate(cbind(NMDS1, NMDS2) ~ Cape, data = scoresmat, FUN = "mean")
sitecent <- aggregate(cbind(NMDS1, NMDS2) ~ Site, data = scoresmat, FUN = "mean")
periodcent <- aggregate(cbind(NMDS1, NMDS2) ~ PeriodMHW, data = scoresmat, FUN = "mean")
```

Create empty NMDS plot.
```{r}
NMDSplot <- ggplot(data = scoresmat, aes(x = NMDS1, y = NMDS2)) +
  theme(axis.title = element_blank(),
        #element_text(size = 10, face = "bold", color = "gray10",
                                 # family = "Roboto") #use for axis titles
        axis.text = element_text(size = 13, color ="gray10", family = "Roboto"),
        axis.ticks = element_line(color = "gray10"),
        panel.background = element_blank(),
        panel.border = element_rect(fill = NA, color = "gray10", linewidth = 1),
        legend.key = element_blank(), 
        legend.position = "right",
        legend.box.spacing = unit(0.1,"cm"),
        legend.title = element_text(size = 10, face = "bold", color = "gray10",
                                    family = "Roboto Medium"),
        legend.text = element_text(size = 9, color = "gray10", family = "Roboto"),
        legend.spacing.y = unit(0.25, 'cm'),
        plot.tag = element_text(size = 20, color="gray10", family = "Roboto Medium"),
        plot.tag.position = c(0.14,0.96)) +
  coord_fixed(ratio = 1, clip = "on",
              xlim = c(-0.85,0.85),
              ylim = c(-0.85,0.85)) +
  scale_x_continuous(breaks = seq(-0.8,0.8,0.4)) +
  scale_y_continuous(breaks = seq(-0.8,0.8,0.4)) +
  guides(color = guide_legend(override.aes = list(alpha = 1), byrow = TRUE, 
                              keyheight = unit(0.5,"cm")))

stresslab <- grobTree(textGrob(
    paste("Stress: ", round(NMDS$stress, digits = 2)), x=0.81, y=0.925,
    gp=gpar(col="black", fontsize=14, fontfamily = "Roboto")))
```

Plot NMDS colors by period.
```{r}
periodNMDS <- NMDSplot + 
  geom_point(data = scoresmat, aes(color = PeriodMHW), size = 2, alpha = 0.2, 
             show.legend = FALSE) +
  #geom_convexhull(data = scoresmat, aes(color = PeriodMHW), size = 0.75, linetype = 2,
  #                alpha = 0, show.legend = FALSE) + #use for convex hull
  stat_ellipse(aes(color = PeriodMHW), size = 1.2, level = 0.95, show.legend = FALSE) +
  geom_point(data = periodcent, aes(color = PeriodMHW), size = 5, alpha = 1,
             show.legend = FALSE) +
  geom_segment(data = sigspp, aes(x = 0, xend=NMDS1, y=0, yend=NMDS2), 
               arrow = arrow(length = unit(0.25, "cm")), color = "grey10", lwd=0.5) +
  geom_segment(data = sigenv, aes(x = 0, xend=NMDS1, y=0, yend=NMDS2), 
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"), lwd=0.9) +
  geom_text_repel(data = sigspp, aes(x=NMDS1, y=NMDS2, label = Abbreviation, point.size = 2),
                  direction = "both", box.padding = 0.3) +
  geom_label_repel(data = sigenv, aes(x=NMDS1, y=NMDS2, label = Abbreviation),
                  alpha = 1, fontface = "bold",
                  segment.color = "black", segment.linetype = "dashed",
                  direction = "both",  max.overlaps = 4, 
                  nudge_x = c(0.55,-1,-0.1), nudge_y = c(-0.35,-0.1,-0.8),
                  box.padding = 0.3, min.segment.length = 0) +
  scale_color_manual(values = periodcolors, name = "Years",
                     labels = c("2006\u20132014: \nPre-SSWD, \nBefore MHW",
                                "2015\u20132016: \nPost-SSWD, \nDuring MHW",
                                "2017\u20132020: \nPost-SSWD, \nAfter MHW")) +
  labs(tag = "j") + 
  annotation_custom(stresslab, xmin = -0.2, xmax = 0.8, ymin = 0.70, ymax = 0.85)

ggsave(periodNMDS,
       filename = paste0("Figures & Tables/NMDS/Period_",
                         scoresmat$Cape[1],"_",scoresmat$Trt[1],"_ellipse2.jpg"), 
       width = 4, height = 4, units = "in", dpi = 800)
```

Plot NMDS colors by cape.
```{r}
capeNMDS <- NMDSplot + 
  geom_point(data = scoresmat, aes(color = Cape), size = 2, alpha = 0.2, 
             show.legend = FALSE) +
  geom_convexhull(data = scoresmat, aes(color = Cape), size = 0.75, linetype = 2,
                  alpha = 0, show.legend = FALSE) +
  geom_point(data = capecent, aes(color = Cape), size = 4, alpha = 1,
             show.legend = FALSE) +
  geom_segment(data = sigspp, aes(x = 0, xend=NMDS1, y=0, yend=NMDS2), 
               arrow = arrow(length = unit(0.25, "cm")), color = "grey10", lwd=0.3) +
  geom_text_repel(data = sigspp, aes(x=NMDS1, y=NMDS2, label = Abbreviation),
                  cex = 3.75, direction = "both", segment.size = 0.25, box.padding = 0.3) +
  scale_color_manual(values = capecolors,
                     labels = c("Foulweather", "Perpetua",
                                "Blanco", "Mendocino")) +
  labs(tag = paste(scoresmat$Year[1]), color = "Cape") + 
  annotation_custom(stresslab)

ggsave(capeNMDS,
       filename = paste0("Figures & Tables/NMDS/Cape_",scoresmat$Year[1],".png"), 
       width = 4, height = 4, units = "in", dpi = 800)
```

Plot NMDS colors by site.
```{r}
NMDSplot + 
  geom_point(data = scoresmat, aes(color = Site), size = 2, alpha = 0.3) +
  geom_convexhull(data = scoresmat, aes(color = Site), linetype = 2, alpha = 0,
               show.legend = FALSE) +
  geom_point(data = sitecent, aes(color = Site), size = 3, alpha = 1, shape = 8,
           show.legend = FALSE) +
  scale_color_manual(values = sitecolors) +
  theme(legend.position = "right", plot.tag.position = c(0.15,0.95)) +
  labs(tag = paste(scoresmat$Year[1]), color = "Site")
```

Plot NMDS colors by treatment.
```{r}
trtNMDS <- NMDSplot + 
  geom_point(data = scoresmat, aes(color = Trt), size = 2, alpha = 0.2,
             show.legend = FALSE) +
  geom_convexhull(data = scoresmat, aes(color = Trt), size = 0.75, linetype = 2,
                  alpha = 0, show.legend = FALSE) +
  geom_point(data = trtcent, aes(color = Trt), size = 4, alpha = 1, 
             show.legend = FALSE) +  
  geom_segment(data = sigspp, aes(x = 0, xend=NMDS1, y=0, yend=NMDS2), 
               arrow = arrow(length = unit(0.25, "cm")), color = "grey10", lwd=0.3) +
  geom_text_repel(data = sigspp, aes(x=NMDS1, y=NMDS2, label = Abbreviation),
                  cex = 3.75, direction = "both", segment.size = 0.25, box.padding = 0.3) +
  scale_color_manual(values = trtcolors, 
                     labels = c("Control", "Recovery",
                                "Macrophyte-only", "Invertebrate-only")) +
  labs(tag = paste(scoresmat$Year[1]), color = "Treatment") + 
  annotation_custom(stresslab)

ggsave(trtNMDS,
       filename = paste0("Figures & Tables/NMDS/Trt_",scoresmat$Year[1],".png"), 
       width = 4, height = 4, units = "in", dpi = 800)
```

### Multivariate Statistics

Calculate permutational multivariate analysis of variance (PERMANOVA).
```{r}
adonis2(sslFG_BC ~ Year * Cape * Site %in% Cape * Trt + Rep %in% Site,
        data = varmat, na.action = na.exclude, permutations = 100, parallel = 2)
```

Calculate pairwise PERMANOVA to compare Bray-Curtis dissimilarities between periods within treatment and cape.
```{r}
setCape <- "CM" #specify cape
setTrt <- "I" #specify treatment
sslFG_BC_subset <- as.matrix(vegdist(
  filter(sslFGmat, row.names(sslFGmat) %in% 
                    filter(dscmat_subset, Trt==setTrt, Cape==setCape)$PlotID),
  method = "bray"))

pairwise.adonis(x=sslFG_BC_subset,
        filter(dscmat_subset, Trt==setTrt, Cape==setCape)$PeriodMHW, perm = 999)
```
